|

### 2023-02-02 paper.Rmd
autogit

---
title: |
  <center>Pre-Trained Denoising Autoencoders Long Short-Term Memory Networks as probabilistic Models for Estimation of Distribution Genetic Programming</center>
  <center>--------------------------------------------------------</center>
  <center>Student: Roman Höhn</center>
  <center>Date of Birth: 1991-04-14</center>
  <center>Place of Birth: Wiesbaden, Hesse</center>
  <center>Student ID: 2712497</center>
  <center>Supervisor: David Wittenberg</center>
  <center>--------------------------------------------------------</center>
  <center>Master Thesis - M.Sc. Business Education</center>
  <center>FB 03: Chair of Business Administration and Computer Science</center>
  <center>Johannes Gutenberg University Mainz</center>
  |
date: "Date of Submission: `r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: true
    extra_dependencies: ["float"]
    fig_caption: yes
    number_sections: true
    toc_depth: 3
    includes:
      in_header: [header_files/settings.tex]
indent: false
fontsize: 11pt
bibliography: ./ref/ref.bib
link-citations: true
fig.width: 4in
fig.height: 3in
csl: csl/association-for-computing-machinery.csl
#csl: csl/harvard-cite-them-right.csl
#csl: csl/springer-science-reviews.csl
include-before: '`\thispagestyle{empty} \newpage{}`{=latex}'
---

@@

The distribution of final fitness scores for pre-trained and regular DAE-GP are visualized as box plots in figure  \@ref(fig:airfoil-distribution-fitness-30gens). The distribution for regular DAE-GP shows to be preferable to that of pre-trained DAE-GP  since it shows a lower median RMSE of `r round(REG_TEST_MED_FF, 3)` for the test set (`r round(PT_TEST_MED_FF, 3)` for pre-trained DAE-GP) as well as less dispersion of the solutions measured by a standard deviation of `r round(REG_TEST_STDD_FF, 3)` (`r round(PT_TEST_STDD_FF, 3)` for pre-trained DAE-GP). For reference, the median final RMSE for the test set of standard GP is at `r round(GP_TEST_MED_FF, 3)` which is larger by a factor of `r round(GP_TEST_MED_FF / PT_TEST_MED_FF, 3)` if compared to pre-trained DAE-GP.

@@

The run-time analysis in this chapter is based on experiments that were all executed on an Intel® Xeon® W-2245 Processor without the use of graphical processing unit (GPU) acceleration in the training of DAE-LSTM networks. Since the framework that was used is written in the ,comparatively slow, python programming language, the absolute run-time measurements are only of limited meaningfulness. Nonetheless, by analyzing the relative run-time differences between DAE-GP and it's pre-trained alternative, the influence of pre-training on DAE-GP run-time can be explored.

@@

```{r airfoil-runtime-noSampling-total, echo=FALSE, fig.cap="Median Runtime excluding Time for Sampling - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_TimeBenchmark/median_runtime_withoutSampling.png")
```
While investigating other reasons for the differences in run-time, I verified that both DAE-GP variants use the exact same number of trainable parameters for each generations DAE-LSTMS models (total number of trainable parameters DAE-GP=`r format(REG_PARAMS, scientific=F)`, Pre-trained DAE-GP=`r format(PT_PARAMS, scientific=F)`). After profiling a single run of DAE-GP and pre-trained DAE-GP my best explanation for the remaining differences in run-time is an inefficiency in my implementation of pre-training. Figure  \@ref(tab:full-run-realWorldSymReg-problems) shows the cumulative time spend for the 20 most expensive function calls. Even though it is difficult to identify the exact reasons for the performance hit for pre-trained DAE-GP, it shows two interesting findings: Firstly, both algorithms spend relatively more time during model sampling (calls to `DAE_LSTM.py:208(sample)`) than during model training (calls to `training.py:1303(fit)`). Secondly, it shows that internal error handling of the baseline neural network library ´keras´ (calls to `traceback_utils.py:138|50(error_handler)`) takes up a comparatively larger amount of run time for pre-trained DAE-GP than for regular DAE-GP.

@@

To gather more confidence in the previous results, the pre-training strategy for DAE-GP was tested for three additional real world symbolic regression problems taken from @machine_learning_repo. Table \@ref(tab:full-run-realWorldSymReg-problems) briefly summarizes the benchmark problems.

>>>>

---
title: |
  <center>Pre-Trained Denoising Autoencoders Long Short-Term Memory Networks as probabilistic Models for Estimation of Distribution Genetic Programming</center>
  <center>--------------------------------------------------------</center>
  <center>Student: Roman Höhn</center>
  <center>Date of Birth: 1991-04-14</center>
  <center>Place of Birth: Wiesbaden, Hesse</center>
  <center>Student ID: 2712497</center>
  <center>Supervisor: David Wittenberg</center>
  <center>--------------------------------------------------------</center>
  <center>Master Thesis - M.Sc. Business Education</center>
  <center>FB 03: Chair of Business Administration and Computer Science</center>
  <center>Johannes Gutenberg University Mainz</center>
  |
date: "Date of Submission: `r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: true
    extra_dependencies: ["float"]
    fig_caption: yes
    number_sections: true
    toc_depth: 3
    includes:
      in_header: [header_files/settings.tex]
indent: false
fontsize: 12pt
bibliography: ./ref/ref.bib
link-citations: true
fig.width: 4in
fig.height: 3in
csl: csl/association-for-computing-machinery.csl
#csl: csl/harvard-cite-them-right.csl
#csl: csl/springer-science-reviews.csl
include-before: '`\thispagestyle{empty} \newpage{}`{=latex}'
---

@@

The distribution of final fitness scores for pre-trained and regular DAE-GP are visualized as box plots in figure  \@ref(fig:airfoil-distribution-fitness-30gens). The distribution for regular DAE-GP shows to be preferable to that of pre-trained DAE-GP  since it has a lower median RMSE of `r round(REG_TEST_MED_FF, 3)` for the test set (`r round(PT_TEST_MED_FF, 3)` for pre-trained DAE-GP) as well as less dispersion of the solutions measured by a standard deviation of `r round(REG_TEST_STDD_FF, 3)` (`r round(PT_TEST_STDD_FF, 3)` for pre-trained DAE-GP). For reference, the median final RMSE for the test set of standard GP is at `r round(GP_TEST_MED_FF, 3)` which is larger by a factor of `r round(GP_TEST_MED_FF / PT_TEST_MED_FF, 3)` if compared to pre-trained DAE-GP.

@@

The run-time analysis in this chapter is based on experiments that were all executed on an Intel® Xeon® W-2245 Processor without the use of graphical processing unit (GPU) acceleration in the training of DAE-LSTM networks. Since the framework that was used is written in the, comparatively slow, python programming language, the absolute run-time measurements are only of limited meaningfulness. Nonetheless, by analyzing the relative run-time differences between DAE-GP and it's pre-trained alternative, the influence of pre-training on DAE-GP run-time can be explored.

@@

```{r airfoil-runtime-noSampling-total, echo=FALSE, fig.cap="Median Runtime excluding Time for Sampling - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_TimeBenchmark/median_runtime_withoutSampling.png")
```
While investigating other reasons for the differences in run-time, I verified that both DAE-GP variants use the exact same number of trainable parameters for each generations DAE-LSTMS models (total number of trainable parameters DAE-GP=`r format(REG_PARAMS, scientific=F)`, Pre-trained DAE-GP=`r format(PT_PARAMS, scientific=F)`). After profiling a single run of DAE-GP and pre-trained DAE-GP my best explanation for the remaining differences in run-time is an inefficiency in my implementation of pre-training. Figure  \@ref(fig:airfoil-profiling) shows the cumulative time spend for the 20 most expensive function calls. Even though it is difficult to identify the exact reasons for the performance hit for pre-trained DAE-GP, it shows two interesting findings: Firstly, both algorithms spend relatively more time during model sampling (calls to `DAE_LSTM.py:208(sample)`) than during model training (calls to `training.py:1303(fit)`). Secondly, it shows that internal error handling of the baseline neural network library `keras` (calls to `traceback_utils.py:138(error_handler)` and `traceback_utils.py:59(error_handler)`) takes up a comparatively larger amount of run time for pre-trained DAE-GP than for regular DAE-GP.

@@

To gather more confidence in the previous results, the pre-training strategy for DAE-GP was applied for three additional real world symbolic regression problems taken from @machine_learning_repo and tested for statistically significant differences in the underlying distributions. Table \@ref(tab:full-run-realWorldSymReg-problems) briefly summarizes the benchmark problems.

### 2023-02-02 paper.tex
autogit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\usepackage[numbers]{natbib}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

The distribution of final fitness scores for pre-trained and regular DAE-GP are visualized as box plots in figure \ref{fig:airfoil-distribution-fitness-30gens}. The distribution for regular DAE-GP shows to be preferable to that of pre-trained DAE-GP since it shows a lower median RMSE of 6.991 for the test set (7.68 for pre-trained DAE-GP) as well as less dispersion of the solutions measured by a standard deviation of 11.989 (12.548 for pre-trained DAE-GP). For reference, the median final RMSE for the test set of standard GP is at 32.024 which is larger by a factor of 4.17 if compared to pre-trained DAE-GP.

@@

While investigating other reasons for the differences in run-time, I verified that both DAE-GP variants use the exact same number of trainable parameters for each generations DAE-LSTMS models (total number of trainable parameters DAE-GP=573473, Pre-trained DAE-GP=573473). After profiling a single run of DAE-GP and pre-trained DAE-GP my best explanation for the remaining differences in run-time is an inefficiency in my implementation of pre-training. Figure \ref{tab:full-run-realWorldSymReg-problems} shows the cumulative time spend for the 20 most expensive function calls. Even though it is difficult to identify the exact reasons for the performance hit for pre-trained DAE-GP, it shows two interesting findings: Firstly, both algorithms spend relatively more time during model sampling (calls to \texttt{DAE\_LSTM.py:208(sample)}) than during model training (calls to \texttt{training.py:1303(fit)}). Secondly, it shows that internal error handling of the baseline neural network library ´keras´ (calls to \texttt{traceback\_utils.py:138\textbar{}50(error\_handler)}) takes up a comparatively larger amount of run time for pre-trained DAE-GP than for regular DAE-GP.

@@

To gather more confidence in the previous results, the pre-training strategy for DAE-GP was tested for three additional real world symbolic regression problems taken from {[}\protect\hyperlink{ref-machine_learning_repo}{2}{]}. Table \ref{tab:full-run-realWorldSymReg-problems} briefly summarizes the benchmark problems.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\usepackage[numbers]{natbib}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

The distribution of final fitness scores for pre-trained and regular DAE-GP are visualized as box plots in figure \ref{fig:airfoil-distribution-fitness-30gens}. The distribution for regular DAE-GP shows to be preferable to that of pre-trained DAE-GP since it has a lower median RMSE of 6.991 for the test set (7.68 for pre-trained DAE-GP) as well as less dispersion of the solutions measured by a standard deviation of 11.989 (12.548 for pre-trained DAE-GP). For reference, the median final RMSE for the test set of standard GP is at 32.024 which is larger by a factor of 4.17 if compared to pre-trained DAE-GP.

@@

While investigating other reasons for the differences in run-time, I verified that both DAE-GP variants use the exact same number of trainable parameters for each generations DAE-LSTMS models (total number of trainable parameters DAE-GP=573473, Pre-trained DAE-GP=573473). After profiling a single run of DAE-GP and pre-trained DAE-GP my best explanation for the remaining differences in run-time is an inefficiency in my implementation of pre-training. Figure \ref{fig:airfoil-profiling} shows the cumulative time spend for the 20 most expensive function calls. Even though it is difficult to identify the exact reasons for the performance hit for pre-trained DAE-GP, it shows two interesting findings: Firstly, both algorithms spend relatively more time during model sampling (calls to \texttt{DAE\_LSTM.py:208(sample)}) than during model training (calls to \texttt{training.py:1303(fit)}). Secondly, it shows that internal error handling of the baseline neural network library \texttt{keras} (calls to \texttt{traceback\_utils.py:138(error\_handler)} and \texttt{traceback\_utils.py:59(error\_handler)}) takes up a comparatively larger amount of run time for pre-trained DAE-GP than for regular DAE-GP.

@@

To gather more confidence in the previous results, the pre-training strategy for DAE-GP was applied for three additional real world symbolic regression problems taken from {[}\protect\hyperlink{ref-machine_learning_repo}{2}{]} and tested for statistically significant differences in the underlying distributions. Table \ref{tab:full-run-realWorldSymReg-problems} briefly summarizes the benchmark problems.
